Higher-Order unification for free!

### 2024-04-12 paper/main.tex
sec 7.2: typo
wip

\newcommand{\Ue}{\ensuremath{\UnifRel_\lambda}\xspace}
\newcommand{\nUe}{\ensuremath{\nUnifRel_\lambda}\xspace}
\newcommand{\Ee}{\ensuremath{\EqualRel_\lambda}\xspace}
\newcommand{\nEe}{\ensuremath{\nEqualRel_\lambda}\xspace}
\newcommand{\llambda}{\ensuremath{\mathcal{L}_\lambda}\xspace}

@@

\newcommand{\lhs}{\ensuremath{\mathrm{lhs}}\xspace}
\newcommand{\rhs}{\ensuremath{\mathrm{rhs}}\xspace}

@@

\newcommand{\progressetaleft}{\emph{progress-$\eta$-left}\xspace}
\begin{definition}[\progressetaleft]\label{def:progressetaleft}
A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
$X$ becomes rigid. Let $y\in\Gamma$, there are two cases:
\begin{enumerate}
  \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
    we unify the $\eta$-expansion of $X$ with $T$, that is we run
    $\lambda x.X \appsep x\Ue{} T$
    \item if $X = \lambda x.t$ we run $X \Ue{} T$.
\end{enumerate}
\end{definition}

\newcommand{\progressetaright}{\emph{progress-$\eta$-right}\xspace}
\begin{definition}[\progressetaright]\label{def:progressetaright} A link
\linketaM{\Gamma}{X}{T} is removed from \linkStore when either 1)
$\testmaybeeta~T$ does not hold (anymore) or 2) by $\eta$-contracting $T$ to
$T'$, $T'$ is a term not starting with the \elpiIn{lam} constructor. In the
first case, $X$ is unified with $T$ and in the second one, $X$ is unified with
$T'$ (under the context $\Gamma$).
\end{definition}

@@

\newcommand{\progressetadedup}{\emph{progress-$\eta$-deduplicate}\xspace}
\begin{definition}[\progressetadedup]\label{def:progressetadedup}
  A link \linketaM{\Gamma}{X_{\vec{s}}}{T} is removed from \linkStore when
  another link \linketaM{\Delta}{X_{\vec{r}}}{T'} is in  \linkStore.
  By \cref{inv:uvaarity} the length of $\vec{s}$ and $\vec{r}$ is the same
  hence we can move the term $T'$ from $\Delta$ to $\Gamma$ by renaming its
  bound variables, i.e. $T'' = T'[\vec{r}/\vec{s}]$.
  We then run $T \Ue{} T''$ (under the context $\Gamma$).
\end{definition}

@@

\begin{theorem}[Fidelity in \maybeeta]
  Given a list of unification problems \foUnifPb, such that 
  $\forall t, t\in\subterm{\foUnifPb} \land t \notin \notllambda$,
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{We also suppose that any higher-order variable is always applied with
  the same number of arguments. This problem is addressed in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}

@@

\begin{proof}[Proof sketch]
  By the definition of \mapdeduplication, any two same \Fo{} variables $X_1, X_2$
  with different arities are related with \linketa. 
  If one of the two variables is instantiated, the corresponding
  \linketa is triggered instantiating the related variable.
  This allows to make unification fail if $X_1$ and $X_2$ are unified with 
  different terms. Finally, since \foUnifPb contains only terms thar are
  either \wellb or \maybeeta, by \cref{lemma:fidelity-maybeeta},
  we can conclude the proof.
\end{proof}

@@

\begin{proof}[Proof sketch]
  Assume we have a \linkbeta, by contradiction, if $m = 0$, then the original
  \Fo{} term has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a
  list of distinct names (i.e. the list \elpiIn{Extra} is empty). This case is
  however captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and
  no \linkbeta is produced which contradicts our initial assumption.
\end{proof}

@@

\begin{proof}[Proof sketch]
  By construction, the lists $s_1\dots s_n$ and $t_1\dots t_m$ are built by splitting
  the list \elpiIn{Ag} from the original term \elpiIn{fapp [fuva A|Ag]}.
  $s_1\dots s_n$ is the longest prefix of the compiled terms in \elpiIn{Ag} which is
  in \llambda. Therefore, by definition of \llambda, $t_1$ must appear  
  in $s_1\dots s_n$, otherwise $s_1\dots s_n$ is not the longest prefix in
  \llambda, or it is a term with a constructor of \elpiIn{tm} as functor.  
\end{proof}

@@

\newcommand{\progBetaLL}{\emph{progress-\llambda-refine}\xspace}
\newcommand{\progBetaRH}{\emph{progress-\llambda-\rhs}\xspace}
% \newcommand{\progBetaDedup}{\emph{progress-\llambda-dedup}\xspace}
\newcommand{\progBetaFail}{\emph{progress-\llambda-fail}\xspace}

Given a \linkbeta $l$ of the form \linkbetaM{\Gamma}{T}{\rhsBeta}, we provide
$4$ different activation rules:

@@

\begin{proof}[Proof sketch]
  Let $l$ a \linkbeta in the store \linkStore. If $l$ is activated by
  \progBetaRH, then it disappears from \linkStore and \elpiIn{progress}
  terminates. Otherwise, the \rhs of $l$ is made by a variable applied to $m$
  arguments. At each activation of \progBetaLL, $l$ is replaced by a new
  \linkbeta $l^1$ having $m-1$ arguments. At the $m^{th}$ iteration, the
  \linkbeta $l^m$ has no more arguments and is removed from \linkStore.
  Note that at the $m^{th}$ iteration, $m$ new \linketa have been added to
  \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
  terminates. Finally \progBetaFail also guarantees termination since 
  it makes \elpiIn{progress} immediately fails.

  % \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  % \linkbeta disappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  % creates each time a new \linketa and replace the old \linkbeta with a new one.
  % This progression can however triggered at most $m$ times, since each new
  % \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  % the \linkbeta is removed from \linkStore which is now filled by $m$ new
  % \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  % terminates if \linkStore is made by only \linketa and therefore
  % \elpiIn{progress} still terminates.
  %\noteE[inline]{is it ok?}
  \noteE[inline]{funziona. per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione). Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
\end{proof}

@@

\begin{theorem}[Fidelity with \linkbeta]
  The introduction of \linkbeta guarantees \fullref{prop:fidelity}
\end{theorem}

\begin{proof}[Proof sketch]
  Let \hoUnifPb a unification problem and $\sigma$ a substitution
  such that $\hoUnifPb \in \notllambda$. 
  If $\sigma\hoUnifPb$ is in \llambda, then by
  \cref{def:progBetaLL,def:progBetaRH}, the \linkbeta associated to the subterm of
  \hoUnifPb have been solved and removed. The
  unification is done between terms in \llambda and by \cref{lemma:fidelity-maybeeta}
  fidelity is guaranteed. If $\sigma\hoUnifPb$ is in \notllambda,
  then, by \cref{def:progBetaLH}, the unification fails, as per the
  corresponding unification in \Fo{}.
\end{proof}

@@

% make test ONLY=7004 TEX=tex
\printAlll
  {{{X,\lambda x.x},
    {\lambda x.(Y\appsep (X\appsep x)),f}}}
  {{{A,\lambda x.x},
    {B,f}}}
  {{{Y,D,0},
    {X,A,0}}}
  {{{\eta,,A,\lambda x.E_{x}},
    {\eta,,B,\lambda x.C_{x}},
    {\beta,x,C_{x},(D\appsep E_{x})}}}

@@

\paragraph{Example of \progBetaRH}
We can take the example provided in \cref{sec:beta}. The problem is compiled
into:
%
% make test ONLY=7003 TEX=tex
\printAlll
  {{{X,\lambda x.Y},
    {(X\appsep a),a}}}
  {{{A,\lambda x.B},
    {C,a}}}
  {{{Y,B,0},
    {X,A,0}}}
  {{{\beta,,C,(A\appsep a)}}}

@@

\section{Conclusion}

>>>>

\newcommand{\Ue}{\ensuremath{\UnifRel_\lambda}\xspace}
\newcommand{\nUe}{\ensuremath{\nUnifRel_\lambda}\xspace}
\newcommand{\Ee}{\ensuremath{\EqualRel_\lambda}\xspace}
\newcommand{\nEe}{\ensuremath{\nEqualRel_\lambda}\xspace}
\newcommand{\llambda}{\ensuremath{\mathcal{L}}\xspace}

@@

\newcommand{\lhs}{lhs\xspace}
\newcommand{\rhs}{rhs\xspace}

@@

\newcommand{\progressetaleft}{\emph{$\eta$-progress-\lhs}\xspace}
\begin{definition}[\progressetaleft]\label{def:progressetaleft}
A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
$X$ becomes rigid. Let $y\in\Gamma$, there are two cases:
\begin{enumerate}
  \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
    we unify the $\eta$-expansion of $X$ with $T$, that is we run
    $\lambda x.X \appsep x\Ue{} T$
    \item if $X = \lambda x.t$ we run $X \Ue{} T$.
\end{enumerate}
\end{definition}

\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
\begin{definition}[\progressetaright]\label{def:progressetaright} A link
\linketaM{\Gamma}{X}{T} is removed from \linkStore when either 1)
$\testmaybeeta~T$ does not hold (anymore) or 2) by $\eta$-contracting $T$ to
$T'$, $T'$ is a term not starting with the \elpiIn{lam} constructor. In the
first case, $X$ is unified with $T$ and in the second one, $X$ is unified with
$T'$ (under the context $\Gamma$).
\end{definition}

@@

\newcommand{\progressetadedup}{\emph{$\eta$-progress-deduplicate}\xspace}
\begin{definition}[\progressetadedup]\label{def:progressetadedup}
  A link \linketaM{\Gamma}{X_{\vec{s}}}{T} is removed from \linkStore when
  another link \linketaM{\Delta}{X_{\vec{r}}}{T'} is in  \linkStore.
  By \cref{inv:uvaarity} the length of $\vec{s}$ and $\vec{r}$ is the same
  hence we can move the term $T'$ from $\Delta$ to $\Gamma$ by renaming its
  bound variables, i.e. $T'' = T'[\vec{r}/\vec{s}]$.
  We then run $T \Ue{} T''$ (under the context $\Gamma$).
\end{definition}

@@

\begin{theorem}[Fidelity in \maybeeta]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb, such that \noteD{riprendere sec 3.1}
  $\forall t, t\in\subterm{\foUnifPb} \land t \notin \notllambda$, \noteD{dire che M e biiettivo}
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{We also suppose that any higher-order variable is always applied with
  the same number of arguments. This problem is addressed in \cref{sec:invariant1}}\noteD{modificare accordingly}
  \label{lemma:fidelity-maybeeta}
\end{theorem}

@@

\begin{proof}[Proof sketch]
  By the definition of \mapdeduplication, any two occurrencies of the same \Fo{} 
  variables $X_1, X_2$
  with different arities are related with \linketa. 
  If one of the two variables is instantiated, the corresponding
  \linketa is triggered instantiating the related variable.
  This allows to make unification fail if $X_1$ and $X_2$ are unified with 
  different terms. Finally, since \foUnifPb contains only terms thar are
  either \wellb or \maybeeta, by \cref{lemma:fidelity-maybeeta},
  we can conclude the proof.
\end{proof}

@@

\begin{comment}
\begin{proof}[Proof sketch]
  Assume we have a \linkbeta, by contradiction, if $m = 0$, then the original
  \Fo{} term has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a
  list of distinct names (i.e. the list \elpiIn{Extra} is empty). This case is
  however captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and
  no \linkbeta is produced which contradicts our initial assumption.
\end{proof}
\end{comment}

@@

\begin{comment}
\begin{proof}[Proof sketch]
  By construction, the lists $s_1\dots s_n$ and $t_1\dots t_m$ are built by splitting
  the list \elpiIn{Ag} from the original term \elpiIn{fapp [fuva A|Ag]}.
  $s_1\dots s_n$ is the longest prefix of the compiled terms in \elpiIn{Ag} which is
  in \llambda. Therefore, by definition of \llambda, $t_1$ must appear  
  in $s_1\dots s_n$, otherwise $s_1\dots s_n$ is not the longest prefix in
  \llambda, or it is a term with a constructor of \elpiIn{tm} as functor.  
\end{proof}
\end{comment}

@@

\newcommand{\progBetaLL}{\emph{\llambda-progress-refine}\xspace}
\newcommand{\progBetaRH}{\emph{\llambda-progress-\rhs}\xspace}
% \newcommand{\progBetaDedup}{\emph{progress-\llambda-dedup}\xspace}
\newcommand{\progBetaFail}{\emph{\llambda-progress-fail}\xspace}

Given a \linkbeta $l$ of the form \linkbetaM{\Gamma}{T}{\rhsBeta}, we provide
$3$ different activation rules:

@@

\begin{proof}[Proof sketch]
  Let $l$ a \linkbeta in the store \linkStore. If $l$ is activated by
  \progBetaRH{}, then it disappears from \linkStore and \elpiIn{progress}
  terminates. Otherwise, the \rhs of $l$ is made by a variable applied to $m$
  arguments. At each activation of \progBetaLL, $l$ is replaced by a new
  \linkbeta $l^1$ having $m-1$ arguments. At the $m^{th}$ iteration, the
  \linkbeta $l^m$ has no more arguments and is removed from \linkStore.
  Note that at the $m^{th}$ iteration, $m$ new \linketa have been added to
  \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
  terminates. Finally \progBetaFail also guarantees termination since 
  it makes \elpiIn{progress} immediately fails.

  % \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  % \linkbeta disappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  % creates each time a new \linketa and replace the old \linkbeta with a new one.
  % This progression can however triggered at most $m$ times, since each new
  % \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  % the \linkbeta is removed from \linkStore which is now filled by $m$ new
  % \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  % terminates if \linkStore is made by only \linketa and therefore
  % \elpiIn{progress} still terminates.
  %\noteE[inline]{is it ok?}
  % \noteE[inline]{funziona. per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione). Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
\end{proof}

@@

\begin{theorem}[Fidelity with \linkbeta]
  The introduction of \linkbeta guarantees \fullref{prop:fidelity-recovery}
\end{theorem}

\begin{proof}[Proof sketch]
  Let \hoUnifPb a unification problem and $\sigma$ a substitution
  such that $\hoUnifPb \in \notllambda$. 
  If $\sigma\hoUnifPb$ is in \llambda, then by
  \cref{def:progBetaLL,def:progBetaRH}, the \linkbeta associated to the subterm of
  \hoUnifPb have been solved and removed. The
  unification is done between terms in \llambda and by \cref{lemma:fidelity-maybeeta}
  fidelity is guaranteed. If $\sigma\hoUnifPb$ is in \notllambda,
  then, by \cref{def:progBetaFail}, the unification fails, as per the
  corresponding unification in \Fo{}.
\end{proof}

@@

% make test ONLY=7004 TEX=tex
\printAlll
  {{{X,\lambda x.x},
    {\lambda x.(Y\appsep (X\appsep x)),f}}}
  {{{A,\lambda x.x},
    {B,f}}}
  {{{Y,D,0},
    {X,A,0}}}
  {{{\eta,,A,\lambda x.E_{x}},
    {\eta,,B,\lambda x.C_{x}},
    {\llambda,x,C_{x},(D\appsep E_{x})}}}

@@

\paragraph{Example of \progBetaRH}
We can take the example provided in \cref{sec:beta}. The problem is compiled
into:
%
% make test ONLY=7003 TEX=tex
\printAlll
  {{{X,\lambda x.Y},
    {(X\appsep a),a}}}
  {{{A,\lambda x.B},
    {C,a}}}
  {{{Y,B,0},
    {X,A,0}}}
  {{{\llambda,,C,(A\appsep a)}}}

% \todo{Rinominare i print dei problemi}

@@

\section{Related work and conclusion}
